The hidden manifold distance for functional data

### 2019-08-22 index.tex
rmarkdown outputs to body

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={The hidden manifold distance for functional data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{The hidden manifold distance for functional data}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\begin{document}
\maketitle

@@

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Functional data analysis is the statistical analysis of smooth
infinite-dimensional curves. The challenge of analyzing
infinite-dimensional objects is ameliorated by the fact that, since
observations are smooth over their domain, the dimensionality of the
curves is only artifically high. Less studied in functional data
analysis is the idea that the curves may actually live in a submanifold
with low intrinsic dimension. Interesting classes of functional manifold
data include the class of probability density functions and the class of
warped curves of a common template function. Since these spaces are not
vector spaces, even basic operations such as addition and subtraction
require special consideration. In this work we set ourselves the task of
calculating the pairwise geodesic distance between functional
observations that, due to noise, live near a non-linear manifold rather
than exactly on the manifold. We will show that good estimation of the
geodesic distance cannot be accomplished by naive methods such as
smoothing followed by a shortest-path algorithm. Instead, we propose a
dedicated technique for pairwise geodesic distance estimation by first
projecting the functional data to the underlying hidden manifold and
then perofrming operations on this hidden manifold. We conclude the work
by showing the implications for many downstream tasks in functional data
analysis, in particular distanced-based functional classification.

\hypertarget{introduction-susan}{%
\section{Introduction {[}Susan{]}}\label{introduction-susan}}

Let \(X_1,\ldots,X_n\) be a sample of \(n\) independant realizations of
a random variable \(X\) that takes value in the Hilbert space
\(L^2([a,b],\mathbb{R})\). We additionally assume that the function
\(X\) belongs to a low-dimensional nonlinear manifold
\(\mathcal{M}\subset L^2([a,b],\mathbb{R})\), in which case a geodesic
distance taking into account the intrinsic structure of \(\mathcal{M}\)
rather than the \(L^2\) distance should be used as a measure of distance
between two functions in \(\mathcal{M}\). As many statistical methods
depend on some measure of distance, downstream benefits of using the
geodesic distance can be expected.

???insert example that shows pairwise geodesic distances give visually
more interesting results than pairwise \(L_2\) distance.???

Given perfect functional observations, i.e.~with no measurement error
and on a very dense domain grid, the estimation of the geodesic distance
can be accomplished straightforwardly using algorithms such as Floyd's.
However, functional data are observed with noise in such a way that the
observed functional data no longer live on a manifold. Though powerful
techniques exist for smoothing discretely observed noisy functional
data, the smoothing is designed for the \(L_2\) recovery of the original
curve, not to ensure the recovered functional versions of the data lie
on or close to the functional manifold \(\mathcal{M}\).

With this challenge in mind, we put forth a technique for the specific
task of estimating pairwise geodesic distances
\(\{ d_\mathcal{M}(X_i,X_j\}_{i>j}\), to be defined when we have access
only to discretely-observed noisy functional observations that possibly
live off the true manifold. Briefly outline the methodology: MDS plus
density ridge are used in a cool way to estimate (consistently?, let's
hope..) pairwise geodesic distances.

\hypertarget{related-work-susan}{%
\subsection{Related Work {[}Susan{]}}\label{related-work-susan}}

(Chen and Muller 2012)

(Lin and Yao 2017)

(Dimeglio et al. 2014)

\hypertarget{setting-susan}{%
\subsection{Setting {[}Susan{]}}\label{setting-susan}}

Let \(X_1,\ldots,X_n\) be independent realizations of a random function
\(X\in\mathcal{M}\subset L^2([a,b],\mathbb{R})\), where \(\mathcal{M}\)
is an unknown manifold. Suppose that each curve \(X_i\) is observed with
measurements errors on a grid \(T_i=(t_{i1},\ldots,t_{iK})\), i.e.~we
observe a sample of \(K\)-dimensional vectors \(Y_1,\ldots,Y_n\) with
\(Y_{ij} = X_i(t_{ij}) + \epsilon_{ij}\), where the random variables
\(\epsilon_{ij}\) are of mean zero and uncorrelated with each other. We
assume that the grids \(T_1,\ldots,T_n\) are dense.

Let \(\tilde X_1,\ldots,\tilde X_n\) denote the functional versions of
the raw data by some smoothing method. Our proposed methodology is
unspecific to the smoothing method employed. For simplicity however,
let's say we employ spline smoothing to recover the functional versions
of the raw data, i.e.\\
\[ \tilde X_i = \arg\min_{f\in C^2[0,1]}\left\{\sum_{j=1}^{K}\left(f(t_{ij})-Y_{ij}\right)^2+\lambda \|\partial^2_tf\|^2_{L^2}\right\}\]
where \(\lambda>0\) is a tuning parameter controlling the smoothness of
\(\tilde X_i\).

\hypertarget{relevant-differential-geometry-concepts-susan}{%
\subsection{Relevant differential geometry concepts
{[}Susan{]}}\label{relevant-differential-geometry-concepts-susan}}

We first rigorously define the concepts of geodesic and geodesic
distance. Following (Lin et al. 2014), let \((\mathcal{M},g)\) be a
\(d\)-dimensional Riemannian manifold, where \(g\) is a Riemannian
metric tensor on \(\mathcal{M}\) which can be used to assign a metric on
the manifold. Specifically, for each point \(p\) on the manifold, the
Riemannian metric tensor \(g\) has an inner product \(g_p\) on the
tangent space \(T_p \mathcal{M}\). The norm of a tangent vector
\(v \in T_p \mathcal{M}\) is defined as \[||v|| = \sqrt{g_p(v,v)}.\] If
\(\gamma: [a,b] \subset \mathbb{R}\to \mathcal{M}\) is a smooth curve,
its length is then defined as
\[ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt .\] The
geodesic distance between two points \(p,q\) on the manifold
\(\mathcal{M}\), based on this metric tensor \(g\), is defined as
\[ d_g(p,q):=\inf \{l(\gamma): \gamma:[a,b] \to \mathcal{M}\text{ piecewise smooth}, \gamma(a) = p, \gamma(b) = q]  \}.\]

\hypertarget{proposed-method-for-estimating-geodesic-distances-marie}{%
\section{Proposed method for estimating geodesic distances
{[}Marie{]}}\label{proposed-method-for-estimating-geodesic-distances-marie}}

We first describe an important helper algorithm which we call IsoGeo. It
is a subroutine of the classic Isomap procedure. Isomap is a three steps
procedure that takes as input a set of points
\(x_1,\ldots,x_n\in \mathbb{R}^D\) and produces an embedding of the
input data in the space \(\mathbb{R}^d\) with \(d<D\), that preserves
pairwise geodesic distances.

\begin{enumerate}
\item The first step consists in constructing a weighted graph $G$ with nodes corresponding to the input data and for which the weight of an edge between two nodes $x_i$ and $x_j$ is equal to $||x_i-x_j ||_{\mathbb{R}^D}$. The graph $G$ is constructed such that a node is connected to a fix number $N$ of neighbors or to all nodes that are at a distance smaller than a given value $\epsilon$.
\item In the second step, the pairwise geodesic distances $d_\mathcal{M}(x_i,x_j)$
are estimated based on $G$. The geodesic distance between two nodes is simply the length of the shortest path in $G$ between these two nodes, which we calculate with the Floyd-Warshall algorithm [@Warshall1962]
<!-- or Dijkstra’s  which one do we actually use in our implementation, the FW or D? -->
\item The last step consists in using MDS on the matrix of pairwise geodesic distances obtained in the first step in order to obtain an embedding in $\mathbb{R}^d$.
\end{enumerate}

In what follows, we call IsoGeo the procedure of estimating pairwise
geodesic distances with the two first steps of Isomap. Note that Floyd
or Dijkstra are methods to find the smallest path given a weigthed graph
but they are not related to how to calculate that graph (which is the
difficult part, depend on number of neighbors, many method out there to
calculate that graph in a robust way, P-ISOMAP is one of these).

Our method is described by the following steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Obtain a \(s\)-dimensional representation
  \(\tilde X^s_1,\ldots,\tilde X^s_n\) of the functions
  \(\tilde X_1,\ldots,\tilde X_n\) that ``preserves'' the pairwise
  \(L^2\) distances by using MDS.
\item
  Obtain
  \(\tilde X^{s,\hat \mathcal{M}}_1,\ldots,\tilde X^{s,\hat \mathcal{M}}_n\)
  a ``projection'' of \(\tilde X^s_1,\ldots,\tilde X^s_n\) onto a ridge
  \(\hat \mathcal{M}\) which is computed with the subspace constrained
  mean-shift algorithm (Ozertem and Erdogmus 2011).
\item
  Use IsoGeo to approximate the pairwise geodesic distances
  \(\{d_{\hat \mathcal{M}}(\tilde X^{s,\hat \mathcal{M}}_i,\tilde X^{s,\hat \mathcal{M}}_j)\}_{i>j}\)
  and define the \(n\) by \(n\) matrix \[
  \hat G(i,j)=\hat G(j,i) = \left\{ \begin{array}{ll}
   d_{\hat \mathcal{M}}(\tilde X^{s,\hat \mathcal{M}}_i,\tilde X^{s,\hat \mathcal{M}}_j) & \textrm{if $i\neq j$,}\\
   0 & \textrm{otherwise.}
    \end{array} \right.
  \]
\end{enumerate}

Since the ridge estimation obtained from noisy measurements of a
manifold should approximate well the manifold (Genovese et al. 2014), we
expect the points
\(\tilde X^{s,\hat \mathcal{M}}_1,\ldots,\tilde X^{s,\hat \mathcal{M}}_n\)
to lie close to the real manifold \(\mathcal{M}\) and then
\(d_{\hat \mathcal{M}}\) to be close to \(d_\mathcal{M}\). Ridge
estimation suffers from the curse of dimensionality, this is why we
first reduce the dimension of our data with MDS and then apply the
shift-mean algorithm to estimate the ridge.

\hypertarget{selection-of-tuning-parameters-marie}{%
\subsection{Selection of tuning parameters
{[}Marie{]}}\label{selection-of-tuning-parameters-marie}}

Describe heuristics. We pick the bandwidth \(h\) in the subspace
constrained mean-shift using Equation (A1) of (Chen et al. 2015).

\hypertarget{simulation-study}{%
\section{Simulation study}\label{simulation-study}}

We perform a simulation study to ascertain the efficacy of our method
for estimating pairwise geodesic distances for discretely-observed noisy
functional data.

\hypertarget{alternative-estimators-of-geodesic-distance-marie}{%
\subsection{Alternative estimators of geodesic distance
{[}Marie{]}}\label{alternative-estimators-of-geodesic-distance-marie}}

\begin{itemize}
\tightlist
\item
  \textbf{(Raw Data) RD} Apply IsoGeo on the raw data \(Y_1,\ldots,Y_n\)
  to obtain an estimator \(\hat G_{\textrm{RD}}\). Note that this
  procedure can only be used if the grid \(T_i\) is the same for each
  \(i\).
\item
  \textbf{(Spline Smoothing) SS} Transform each vector \(Y_i\) into a
  function \(\tilde X_i\) by spline smoothing. Apply IsoGeo on the
  vectors \(\{(\tilde X_i(t_1), \ldots, X_i(t_K)\}_{i=1}^n\), where
  \(t_1,\ldots,t_K\) is a regular grid of \([a,b]\), to obtain an
  estimator \(\hat G_{\textrm{SS}}\).
\item
  \textbf{(P-ISOMAP) pI} Transform each vector \(Y_i\) into a function
  \(\tilde X_i\) by spline smoothing. Calculate the weighted graph of
  step I of Isomap with a penalty (describe the penalty of Chen and
  Muller). Apply step II of Isomap to obtain \(\hat G_{\textrm{pI}}\).
\item
  \textbf{(Random Projection) RP} same method as our but change step 2
  by : obtain \(s\)-dimensional representation by random projection and
  setp 4 by obtain \(\hat G\) using a ensemble method.
\end{itemize}

We ran Chen and Muller's P-ISOMAP separately in Matlab with their
automatic penalty selection and discovered that ???.

\hypertarget{performance-metrics-susan}{%
\subsection{Performance metrics
{[}Susan{]}}\label{performance-metrics-susan}}

We describe three different metrics we use to assess the quality of a
pairwise geodesic distance estimator. This is the ROC curve with
\(\epsilon\) on the \(x\)-axis and degree to which near-\(\epsilon\)
isometry holds, i.e.~the percentage of estimated pairwise distances
between \(1-\epsilon\) and \(1+\epsilon\) of the truth pairwise
distance.

\hypertarget{simulation-scenarios-susan}{%
\subsection{Simulation scenarios
{[}Susan{]}}\label{simulation-scenarios-susan}}

Acknowledge that we are sampling on a very concentrated measure to make
sure functional manifold is well sampled. Cite Diaconis et. al's
``Sampling from a manifold''. How to sample properly from a functional
manifold could be interesting future work.

\begin{itemize}
\tightlist
\item
  warping manifold (scenario 5 in sim\_functional\_data.R)
\item
  pdf manifold (scenario 2 in sim\_functional\_data.R)
\item
  square root pdf manifold (scenario 4 )
\end{itemize}

\hypertarget{results}{%
\subsubsection{Results}\label{results}}

Submit compare\_methods.sh as slurm job using sbatch.

\begin{Shaded}
\begin{Highlighting}[]
\ExtensionTok{sbatch}\NormalTok{ compare_methods.sh}
\end{Highlighting}
\end{Shaded}

Alternatively, we can run the R script compare\_methods.R locally by
removing the slurm\_arrayid chunk

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{R compare_methods.R}
\end{Highlighting}
\end{Shaded}

Simulation results are then visualized by invoking cluster\_results.R.
There's some unfortunate hardcoding still in cluster\_results.R

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{R cluster_results.R}
\end{Highlighting}
\end{Shaded}

\hypertarget{the-cost-of-employing-proposed-method-when-manifold-is-flat-susanmarie}{%
\subsection{The cost of employing proposed method when manifold is flat
{[}Susan/Marie?{]}}\label{the-cost-of-employing-proposed-method-when-manifold-is-flat-susanmarie}}

Suppose we take scenario 1 where the geodesic distance coincides with
the \(L_2\) distance. Then what is the cost of employing our distance
estimator compared to smoothing the data and doing numerical integration
to find the \(L_2\) distance?

\hypertarget{distance-based-functional-classification-marie}{%
\section{Distance-based functional classification
{[}Marie{]}}\label{distance-based-functional-classification-marie}}

>>>>

@@

\section{Abstract}\label{abstract}

The clustering and classification of functional data, discretely-sampled
data varying over a continuum, is frequently encountered in machine
learning. While one can naively treat the data as vectors in Euclidean
space, one may also regard each data point as a function. The challenge
of analyzing infinite-dimensional functions can be ameliorated if the
curves are smooth since the dimensionality is then only artificially
high. This insight underpins functional data analysis, a subfield of
statistics that studies functional data. For certain types of functional
data such as classes of probability density functions and classes of
warped curves of a common template function, there is further structure
to exploit. Namely, when functional data lie on a manifold, the
dimensionality at play might be even lower if the manifold is itself of
low intrinsic dimension. This work addresses the estimation of pairwise
geodesic distances between functional manifold data that are observed
with noise and which may live near, rather than on, a manifold. The
proposed methodology first sends the observed functional data to the
hidden manifold, estimated using subspace-constrained mean-shift.
Geodesic distances are subsequently calculated by employing
shortest-path algorithms on this estimated manifold. Improved estimation
of the pairwise geodesic distance has beneficial implications for many
downstream tasks in functional data analysis which we illustrate in the
case of distance-based functional classification.

\section{Introduction}\label{introduction}

Many statistical and machine learning methods rely on some measure of
distance. For example, clustering and classification of functional data,
discretely-sampled data varying over a continuum, is a common machine
learning task in which distance plays a crucial role. The successful
development of functional data analysis, a branch of statistics
dedicated to the analysis of infinite-dimensional smooth curves,
demonstrates that employing the \(L^2\) distance between the smoothed
functional observations has advantages over the Euclidean distance
between functional observations treated as vectors. In this work, we
advocate that for functional manifold data, i.e.~functional data that
lie on a manifold, the geodesic distance is potentially even better than
the \(L^2\) distance, with benefits realized in downstream tasks such as
distanced-based clustering and classification of functional manifold
data. In other words, when the manifold hypothesis is plausible,
e.g.~classes of probability density functions and classes of warped
curves of a common template function, clustering and classification
using the goedesic distance may give better results than using the
\(L_2\) distance.

Let \(X_1,\ldots,X_n\) be a sample of \(n\) independant realizations of
a random variable \(X\) that takes value in the Hilbert space
\(L^2([a,b],\R)\). Suppose additionally that the function \(X\) belongs
to a Riemannian manifold \(\M \subset L^2([a,b],\R)\) with Riemannian
metric tensor \(g\) which can be used to assign a metric on the manifold
as follows (Lin et al. 2014). For each point \(X\) on the manifold, the
Riemannian metric tensor \(g\) has an inner product \(g_X\) on the
tangent space \(T_X \M\). The norm of a tangent vector \(V \in T_X \M\)
is defined as \[||V|| = \sqrt{g_X(V,V)}.\] The geodesic distance between
two functions \(X_i,X_j\) on the manifold \(\M\), based on this metric
tensor \(g\), is defined as
\[ d_{\M}(X_i,X_j):=\inf \{l(\gamma): \gamma \text{ is piecewise-smooth function from } [a,b] \text{ to } \M, \gamma(a) = X_i, \gamma(b) = X_j \}\]
where \[ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt\]
is the length of the curve \(\gamma\). ???Is it clear how \(l(\gamma)\)
depends on \(g\)???

Ideally, the functions \(X_i\) and \(X_j\) are observed on a very dense
domain grid with no measurement error. Then the estimation of the
geodesic distance \(d_\M(X_i,X_j)\) is straightfoward, e.g.~a
shortest-path algorithm such as Floyd-Warhsall (Floyd 1962) may be used.
However, most shortest-path algorithms critically assume observations
live on the manifold ???add reference??? and are thus likely to fail for
functional data which are almost certainly observed with noise. In this
work, we put forth a technique for the recovery of the \(n{\times}n\)
matrix \(G\) of pairwise geodesic distances, i.e.\\
\[
G(i,j)=G(j,i) = \begin{cases} 
      d_{\M} (X_i,X_j) & \textrm{if $i\neq j$,} \\
      0 & \textrm{otherwise.}
   \end{cases}
\] In lieu of \(X_i\) and \(X_j\), we have access only to
discretely-sampled noisy functional observations \(Y_i\) and \(Y_j\)
that possibly lie off the true manifold.

The work of (Chen and Muller 2012) was among the first in functional
data analysis to consider the manifold hypothesis for functional data. A
notion of the mean and variation of functional manifold data was
introduced there. Of particular relevance to this work is their proposed
P-ISOMAP procedure which adds a penalty to allow for noisy functional
observations whereas the classic ISOMAP algorithm (Tenenbaum, Silva, and
Langford 2000) assumes observations lie exactly on the manifold.
(Dimeglio et al. 2014) also realized this drawback to ISOMAP and
proposed a procedure we will call robust-ISOMAP that is less sensitive
to outliers. We will discuss P-ISOMAP and robust-ISOMAP in more details
in the simulation section.

\section{Proposed method for estimating geodesic
distances}\label{proposed-method-for-estimating-geodesic-distances}

Suppose that each curve \(X_i \in (\M,g)\) is observed with measurement
error on a time grid
\(T_i=(t_{i1},\ldots,t_{iK}), a \le t_{i1} < \ldots < t_{iK} \le b\),
i.e.~we observe a sample of \(K\)-dimensional vectors \(Y_1,\ldots,Y_n\)
with \(Y_{ij} = X_i(t_{ij}) + \epsilon_{ij}\), where the random
variables \(\epsilon_{ij}\) have mean zero and are uncorrelated with
each other. We assume that each observation grid \(T_1,\ldots,T_n\) is
dense, i.e.~\(K\) is large.

We begin by converting the discretly-sampled functional observations
into continuous ones. Let \(\tilde X_1,\ldots,\tilde X_n\) denote the
functional versions of the raw data obtained by some smoothing method.
For example, we may employ spline smoothing ???add reference for spline
smoothing??? to recover the functional versions of the raw data, i.e.\\

\begin{equation}\label{eq_spline_smoothing} 
\tilde X_i = \arg \min_{f\in C^2[0,1]}\left\{\sum_{j=1}^{K}\left(f(t_{ij})-Y_{ij}\right)^2+\lambda \|\partial^2_tf\|^2_{L^2}\right\},
\end{equation}

where \(\lambda>0\) is a tuning parameter controlling the smoothness of
\(\tilde X_i\). The proposed methodology for estimating the pairwise
geodesic distances \(\{ d_\M (X_i,X_j)\}_{i>j}\) is agnostic to the
smoothing method employed.

Our method is based on the idea that the underlying functional manifold
\(\M\) can be sufficiently well-recovered by the subspace-constrained
mean-shift (SCMS) algorithm (Ozertem and Erdogmus 2011). Let
\(\hat M \subset L^2([a,b],\mathbb R)\) denote the collection of basins
of attraction of the estimated probability density function of \(X\).
The SCMS algorithm sends each point
\(\tilde X_i \in L^2([a,b],\mathbb R)\) to a point in \(\hat M\). This
destination is unique and we denote it \(\tilde X_i^{\hat \M}\). We
expect the points
\(\tilde X^{\hat \M}_1,\ldots,\tilde X^{\hat \M}_n \in \hat M\) to lie
close to the real manifold \(\M\). Theoretical justificaton that
\(\hat M\) as estimated by SCMS is a reasonable surrogate for the true
manifold \(\M\) can be found in (Genovese et al. 2014).

Before presenting our procedure, we need to first review the ISOMAP
algorithm (Tenenbaum, Silva, and Langford 2000), a three-step procedure
that takes as input a set of points \(x_1,\ldots,x_n\) in a submanifold
\(\M\) of \(\R^D\) and produces an embedding in the space \(\R^d\) with
\(d<D\) that preserves pairwise geodesic distances. The procedure is as
follows.

\begin{enumerate}
\item Construct a weighted graph $G$ with nodes corresponding to the observations $x_1,\ldots,x_n\in \R^D$. Two nodes $x_i$ and $x_j$ are connected by an edge $e_{ij}$ with weight $d_{ij} 1\{d_{ij} \le \epsilon \}$ where $d_{ij}=||x_i-x_j ||_{2}$ and $\epsilon > 0$.
\item Estimate the pairwise geodesic distances $d_\M(x_i,x_j)$ based on $G$ using shortest-path algorithms. Specifically, the geodesic distance between two nodes is estimated to be the length of the shortest path in $G$ between these two nodes, i.e.\ the sum of the weights of the edges forming the shortest path, which is calculated either with the Floyd-Warshall algorithm [@Floyd1962] or with the Dijkstra algorithm [@Dijkstra59anote].
\item Use multidimensional scaling ???add reference??? to obtain an embedding in $\R^d$ that preserves the pairwise geodesic distances estimated above.
\end{enumerate}

Note that shortest-path algorithms such as the Floyd-Warshall algorithm
are used to find the smallest path \textit{given} a weigthed graph but
they do not produce the weighted graphs themselves. Both P-ISOMAP and
robust-ISOMAP are modifications of ISOMAP in the sense that they change
the construction of the weighted graph in step 1.\\

In what follows, we call IsoGeo the two-step procedure which performs a
modified version of the first step of ISOMAP followed by the original
second step of ISOMAP. The first step of IsoGeo constructs the weighted
graph \(G\) accoding to Dimeglio et al. (2014) so that graph is
independent of the tuning parameter \(\epsilon\). Specifically, we begin
by constructing the complete ???explain what complete means??? weighted
graph \(G_c\) with nodes corresponding to the observations
\(x_1,\ldots,x_n\in \R^D\). Next, we obtain the minimal spanning tree
associated with \(G_c\) and denote its set of edges by \(E_s\). The
graph \(G\) is obtained by adding all edges \(e_{ij}\) to \(E_s\) for
which the following condition is true:
\[ \overline{x_ix_j} \subset \bigcup_{i=1}^n B(x_i,\epsilon_i),\] where
\(B(x_i,\epsilon_i)\) is the open ball of center \(x_i\) and radius
\(\epsilon_i = \max_{e_{ij}\in E_s}d_{ij}\), and
\(\overline{x_ix_j}= \{x\in R^D \ | \ \exists \lambda \in [0,1], x=\lambda x_i + (1-\lambda)x_j \}\).
Let the distance estimated by IsoGeo be denoted
\(\text{IsoGeo}(x_i,x_j)\). Note that IsoGeo has no tuning parameters.

We are now ready to describe our procedure. Since SCMS is based on
kernel density estimation, we first reduce the dimension of our data
with multidimensional scaling before applying SCMS to avoid the curse of
dimensionality.

\begin{enumerate}
\item Discretise the functions $\tilde X_1,\ldots,\tilde X_n$ by evaluating them on a dense common grid $\tilde T=(t_{1},\ldots,t_{D})$. Let $s$ be a positive integer, much smaller than $D$. Obtain $\tilde X^s_1,\ldots,\tilde X^s_n \in \mathbb R^s$ using multidimensional scaling whereby the pairwise $\mathbb R^D$ Euclidean distances of the discretised functions $\tilde X_i$ are preserved.
\item Apply the subspace constrained mean-shift algorithm [@Ozertem2011] to each of $\tilde X^s_i$ to obtain $\tilde X^{s,\hat \M}_i$.
\item Use IsoGeo on the inputs $(\tilde X^{s,\hat \M}_1,\ldots,\tilde X^{s,\hat \M}_n)$ to obtain $\text{IsoGeo}(\tilde X^{s,\hat \M}_i,\tilde X^{s,\hat \M}_j)\}_{i>j}$.
\end{enumerate}

Our geodesic distance estimator is the \(n{\times}n\) matrix \(\hat G\)
whose elements are given by \[
\hat G(i,j)=\hat G(j,i) = \left\{ \begin{array}{ll}
 \text{IsoGeo}(\tilde X^{s,\hat \M}_i,\tilde X^{s,\hat \M}_j) & \textrm{if $i\neq j$,}\\
 0 & \textrm{otherwise.}
  \end{array} \right.
\]

There are two tuning parameters to our procedure. First, there is the
dimension \(s\) in step 1. In our simulations we try
\(s \in \{1,2,3\}\). The other parameter is the bandwidth \(h\) in the
subspace constrained mean-shift. In our simulations we set \(h\)
according to the heuristic in Equation (A1) of (Chen et al. 2015). In
reality, depending on the downstream task, both \(s\) and \(h\) can be
tuned in a data-adaptive way, say using cross-validation.

\section{Simulation study}\label{simulation-study}

We perform a simulation study to ascertain the efficacy of our method
for estimating pairwise geodesic distances. Three different metrics are
used to assess the quality of a pairwise geodesic distance estimator.

The first metric assesses near-\(\epsilon\) isometry, which we define as
the percentage of estimated pairwise geodesic distances between
\(1-\epsilon\) and \(1+\epsilon\) of the corresponding true pairwise
geodesic distance. We form the receiver operating curve with
\(\epsilon\) on the \(x\)-axis and near-\(\epsilon\) isometry on the
\(y\)-axis. The metric is the area under the receiver operating curve.

The second metric \(||d-\hat d||/||d||\) is the Frobenius norm of the
estimation error matrix \(d-\hat d\), relative to the Frobenius norm of
the true distance matrix \(d\). The third metric is the Pearson
correlation coefficient between the upper diagonal of \(d\) and
\(\hat d\).

\subsection{Alternative estimators of geodesic distance
{[}Marie{]}}\label{alternative-estimators-of-geodesic-distance-marie}

We compare the performance of our method to each of the following four
methods. The first method, denoted RD for raw data, takes the naive
approach of applying IsoGeo directly on the raw vectors
\(Y_1,\ldots,Y_n \in R^K\). Note that this procedure is only sensible if
all the grids \(T_1,\ldots,T_n\) are the same. In the second method, we
first smooth the discretely-sampled noisy functional data and then
estimate their pairwise distance using the \(L^2\) distance.

The third alternative, denoted SS for spline smoothing, applies IsoGeo
on the smoothed version \(\tilde X_1,\ldots,\tilde X_n\) of the raw data
obtained by spline smoothing as described in
(\ref{eq_spline_smoothing}). The final alternative we consider is
P-ISOMAP, denoted pI, which is the two-step procedure developed in Chen
and Muller (2012) designed to handle noisy functional data. In the first
step, a penalty is incorporated to robustify the construction of the
weighted graph. This is followed by the original second step of ISOMAP.

We ran Chen and Muller's P-ISOMAP separately in Matlab and tuned for the
number of neighbors and the penalty parameter using the relative
Frobenius assessment metric. We found that the penalty parameter chosen
in this way was always zero, thus reducing hte P-ISOMAP procedure to
standard ISOMAP. Given this, we did not make further comparison to
P-ISOMAP in our simulation study.

\subsection{Simulation scenarios}\label{simulation-scenarios}
